# Confidence intervals


You have run your experiment and presented your readers with your single best guess about the treatment effect with the difference in sample means. You may have also presented the estimated standard error of this estimate to give readers a sense of how variable the estimate is. You might have even done a hypothesis test against the null of no treatment effect. But none of these approaches answer a fairly compelling question: what range of values of the treatment effect are **plausible** given the data we observe?

A point estimate typically has 0 probability of being the exact true value, but intuitively we hope that the true value is close to this estimate. **Confidence intervals** make this kind of intuition more formal by instead estimating ranges of values with a fixed percentage of these ranges containing the true value of the parameter. 

We begin with the basic definition of a confidence interval. 

::: {#def-coverage}
A $1-\alpha$ **confidence interval** for a real-valued parameter $\theta$ is a pair of statistics $L= L(X_1, \ldots, X_n)$ and $U = U(X_1, \ldots, X_n)$ such that $L < U$ for all values of the sample and such that 
$$ 
\P(L \leq \theta \leq U \mid \theta) \geq 1-\alpha, \quad \forall \theta \in \Theta.
$$
:::

We say that that a $1-\alpha$ confidence intervals covers (contains, captures, traps, etc) the true value at least $100(1-\alpha)\%$ of the time, and we refer to $1-\alpha$ as the **coverage probability** or simply **coverage**. Common confidence intervals include 95% percent ($\alpha = 0.05$) and 90% ($\alpha = 0.1$). 

So a confidence intervals is a random interval that has a particular guarantee about how often it will contain the true value. It's important to remember what is random and what is fixed in this setup. The interval varies from sample to sample but the true value of the parameter stays fixed, and the coverage is how often we should expect the interval to contain that true value. The "repeating my sample over and over again" analogy can break down very quickly, so it's sometimes helpful to interpret as giving guarantees across confidence intervals across different experiments. In particular, suppose that a journal publishes 100 quantitative articles per year and each produces a single 95% confidence interval for their quantity of interest. Then, if the confidence intervals are valid, then we should expect 5 of those confidence intervals to not contain their true quantity of interest. 


::: {.callout-warning}

Let's say you calculate a 95% confidence interval and it's $[0.1, 0.4]$. It's tempting to make a probability statement like $\P(0.1 \leq \theta \leq 0.4 \mid \theta) = 0.95$ or that there's a 95% chance that the true parameter is in $[0.1, 0.4]$. But looking at the probability statement, everything on left-hand side of the conditioning bar is fixed, so the probability either has to be 0 ($\theta$ is outside the interval) or 1 ($\theta$ is in the interval). The coverage probability of a confidence interval refers to its status as a pair of random variables, $(L, U)$, not any particular realization of those variables like $(0.1, 0.4)$. As an analogy, considering if calculated the sample mean as $0.25$ and then tried to say that $0.25$ is unbiased for the population mean. This doesn't make sense because unbiasedness refers to how the sample mean varies from sample to sample. 

:::


In most cases, we will not be able to derive exact confidence intervals, but rather confidence intervals that are **asymptotically valid**, which means that if we write the interval as a function of the sample size, $(L_n, U_n)$ they would have **asymptotic coverage**
$$
\lim_{n\to\infty} \P(L_n \leq \theta \leq U_n) \geq 1-\alpha \quad\forall\theta\in\Theta.
$$

This will usually be the property that we can show for most confidence intervals, since we usually rely on large sample approximations based on the central limit theorem. 



## Deriving confidence intervals {#sec-deriving-cis}

A thought experiment can give the basic intuition for almost all approaches to generating confidence intervals. Suppose we've estimated support for increasing legal immigration to be $\widehat{\theta}_n = 0.56$ from a sample of size $n = 400$. Our goal is to derive a probabilistic bound on the true proportion from this sample. We can start with trying to derive the probablistic lower bound. 


Recall that we can think of all estimators as the true value plus some estimation error,
$$
\widehat{\theta}_n = \theta + \text{estimation error}.
$$
We already know the estimate, so if we knew the estimation error, then we would could back out the true value of the parameter, $\theta$. Obviously, we don't know the estimation error, but in many cases we know (or can approximate) the *distribution* of the estimation error. For example, for asymptotically normal estimators, we know that, in large samples,
$$
\widehat{\theta}_n - \theta \sim N(0, \se[\widehat{\theta}_n]).
$$
In the case of the proportion support for increasing legal immigration, we know that $\se[\widehat{\theta}_n] = \sqrt{\theta(1 - \theta)/n}$ given that the outcome is binary.  

From this insight, we can engage our thought experiment: what if we had an extremely large positive estimation error that was, say, larger than 97.5% of estimation errors?  Intuitively, if we have a large estimator error, then the true value of $\theta$ should be much smaller than our estimate and could provide a probabilistic lower bound for $\theta$. From the normal distribution, this estimation error would be 
$$
\widehat{\theta}_n - \theta = 1.96 \times \se[\widehat{\theta}_n]
$$
from which we can back out the implied true value of the parameter,
$$
L = \widehat{\theta}_n - 1.96 \times \se[\widehat{\theta}_n].
$$
Here, $L$ is a probabilistic or random lower bound because the value $\widehat{\theta}_n$ is random. Sometimes it will cover the true value ($L \leq \theta$) and sometimes it will not cover the true value ($L > \theta$). We controlled how often it covers the true value. 
Because 2.5% of estimation errors will be larger than $1.96 \times \se[\widehat{\theta}_n]$, then $L$ will be a valid lower bound (cover the true value) in 97.5% of samples, or $\P(L \leq \theta) \approx 0.975$ and not cover the true value 2.5% of samples. 

In the immigration esample, we have 
$$
L = 0.56 - 1.96 \times \sqrt{0.56(1 - 0.56)/400} \approx 0.56 - 0.048 = 0.511,
$$
where we have used the usual plugin estimator for the standard error. A lower bound of $0.511$ tells us that values below that are only compatible with very large and rare chance errors. What if a 2.5% chance of this bound being wrong is too high? We could instead consider what the truth would be for a positive chance error that is larger than 99% of chance errors, which would be $z_{0.99} \times \se[\widehat{\theta}_n] = 2.32 \times \se[\widehat{\theta}_n]$. In the immimgration example, we have
$$
L = 0.56 - 2.32 \times \sqrt{0.56(1 - 0.56)/400} \approx 0.502,
$$
which is a lower lower bound. This refelcts a fundamental tradeoff: our probabilistic bounds get worse (rule out a smaller region of possible values) as we base them off of more extreme chance errors. 

A similar approach can help to derive an upper bound. We imagine a very large negative estimation error that implies a probabilistic upper bound on $\theta$. A very large negative estimation error could be $-1.96 \times \se[\widehat{\theta}_n]$ which is smaller than 2.5% of estimation errors. This result would imply a probabilistic upper bound of
$$
U = \widehat{\theta}_n + 1.96 \times \se[\widehat{\theta}_n],
$$
which will not cover the true value 2.5% of the time, $\P(U < \theta) \approx 0.025$. 

We have derived a probabilistic upper bound and lower bound separately, but we can easily combine them into one interval. Intuitively, if our upper bound misses the true value in 2.5% of samples and the lower bound misses the true value in 2.5% of samples, then the interval $[L, U]$ should not cover the true value 5% of the time. Formally, we have:
$$
\P(L \leq \theta \leq U) = 1 - \left(\P(L > \theta) + \P(U < \theta)\right) = 1 - (0.025 + 0.025) = 0.95.
$$

More generally, for a $100(1-\alpha)$% confidence interval, we need to determine the range of chance errors that puts $\alpha/2$ in each of the tails. 

```{r}
#| label: fig-chance-errors
#| echo: false
#| fig-cap: "Chance error distribution."
curve(dnorm(x), from = -4, to = 4, ylim = c(-.05, 0.5), bty = "l", las = 1, xlab = expression(hat(theta)[n] - theta), ylab = expression(f(z)), xaxt = "n")
abline(h = 0)
axis(side = 1, at = c(-1.96, 1.96), labels = c(expression(z[alpha/2] %*% "se"(hat(theta)[n])), expression(z[1-alpha/2]%*% "se"(hat(theta)[n]))))
polygon(c(-4,seq(-4,-1.96,0.01),-1.96), c(0,dnorm(seq(-4,-1.96,0.01)), 0), col = "grey70", border = NA)
polygon(c(4,seq(4,1.96,-0.01),1.96), c(0,dnorm(seq(4,1.96,-0.01)), 0), col = "grey70", border = NA)
abline(v = 1.96, lty = 2, col = "grey60")
abline(v = -1.96, lty = 2, col = "grey60")
## text(x = 1.94, y = -0.04, "c", pos = 4)
## text(x = -1.94, y = -0.04, "-c", pos = 2)
text(x = 0, y = dnorm(0)/2, expression(1-alpha))
text(x = 2.2, y = 0.05, expression(alpha/2), pos = 4)
text(x = -2.2, y = 0.05, expression(alpha/2), pos = 2)
```

Thus, for the $100(1-\alpha)$% confidence interval, we can derive the implied the lower bound as 
$$
L = \widehat{\theta}_n - z_{1-\alpha/2} \se[\widehat{\theta}_n],
$$
and the upper bound as 
$$
U = \widehat{\theta}_n - z_{\alpha/2} \se[\widehat{\theta}_n].
$$
It's slightly unintuitive that we subtract the chance imbalances in both of these expressions, but recall that $z_{\alpha/2}\se[\widehat{\theta}_n]$ will be a negative number (it is, after all, a large negative chance error). In particular, for asymptotically normal estimators, we know that the chance error distribution is symmetric so that $z_{\alpha/2} = -z_{1-\alpha/2}$, leading to the usual confidence interval formula,
$$
[L, U] = \left[\widehat{\theta}_n - z_{1-\alpha/2} \se[\widehat{\theta}_n], \widehat{\theta}_n + z_{1-\alpha/2} \se[\widehat{\theta}_n]\right].
$$

This is sometimes shortened to $\widehat{\theta}_n \pm z_{\alpha/2} \widehat{\se}[\widehat{\theta}_{n}]$. Remember that we can obtain the values of $z_{\alpha/2}$ easily from R:

```{r}
## alpha = 0.1 for 90% CI
qnorm(0.1 / 2, lower.tail = FALSE)
```

As a concrete example, then, we could derive a 90% asymptotic confidence interval for the sample mean as 
$$
\left[\Xbar_{n} - 1.64 \frac{\widehat{\sigma}}{\sqrt{n}}, \Xbar_{n} + 1.64 \frac{\widehat{\sigma}}{\sqrt{n}}\right]
$$




## Deriving confidence intervals (alternate)

If you have taken any statistics before, you probably have seen the standard formula for the 95% confidence interval of the sample mean, 
$$ 
\left[\Xbar_n - 1.96\frac{s}{\sqrt{n}},\; \Xbar_n + 1.96\frac{s}{\sqrt{n}}\right],
$$
where you can recall that $s$ is the sample standard deviation and $s/\sqrt{n}$ is the estimate of the standard error of the sample mean. If this is a 95% confidence interval, then the probability that it contains the population mean $\mu$ should be 0.95, but how can derive this? It turns out that this will be justified by the central limit theorem and we will show the logic for a generic asymptotically normal estimator. 


Let's say that we have an estimator, $\widehat{\theta}_n$ for the parameter $\theta$ with estimated standard error $\widehat{\se}[\widehat{\theta}_n]$. If the estimator is asymptotically normal, then in large samples, we know that
$$ 
\frac{\widehat{\theta}_n - \theta}{\widehat{\se}[\widehat{\theta}_n]} \sim \N(0, 1).
$$
We can then use our knowledge of the standard normal and the empirical rule to find
$$ 
\P\left( -1.96 \leq \frac{\widehat{\theta}_n - \theta}{\widehat{\se}[\widehat{\theta}_n]} \leq 1.96\right) = 0.95
$$
and by multiplying each part of the inequality by $\widehat{\se}[\widehat{\theta}_n]$, we get
$$ 
\P\left( -1.96\,\widehat{\se}[\widehat{\theta}_n] \leq \widehat{\theta}_n - \theta \leq 1.96\,\widehat{\se}[\widehat{\theta}_n]\right) = 0.95,
$$
We then subtract all part by the estimator to get
$$ 
\P\left(-\widehat{\theta}_n - 1.96\,\widehat{\se}[\widehat{\theta}_n] \leq - \theta \leq -\widehat{\theta}_n + 1.96\,\widehat{\se}[\widehat{\theta}_n]\right) = 0.95,
$$
and finally we multiply all parts by $-1$ (and flipping the inequalities) to arrive at
$$ 
\P\left(\widehat{\theta}_n - 1.96\,\widehat{\se}[\widehat{\theta}_n] \leq \theta \leq \widehat{\theta}_n + 1.96\,\widehat{\se}[\widehat{\theta}_n]\right) = 0.95.
$$
To connect back to the definition of the confidence interval, we have now shown that the random interval $[L, U]$ where
$$ 
\begin{aligned}
  L = L(X_1, \ldots, X_n) &= \widehat{\theta}_n - 1.96\,\widehat{\se}[\widehat{\theta}_n] \\
  U = U(X_1, \ldots, X_n) &= \widehat{\theta}_n + 1.96\,\widehat{\se}[\widehat{\theta}_n],
\end{aligned}
$$
is an asymptotically valid estimator.[^1] Replacing $\Xbar_n$ for $\widehat{\theta}_n$ and $s/\sqrt{n}$ for $\widehat{\se}[\widehat{\theta}_n]$ establishes how the standard 95% confidence interval for the sample mean above asymptotically valid. 

[^1]: Implicit in this analysis is that the estimate of the standard error is consistent. 


```{r}
#| label: fig-std-normal
#| echo: false
#| fig-cap: "Critical values for the standard normal."
curve(dnorm(x), from = -4, to = 4, ylim = c(-.05, 0.5), bty = "l", las = 1, xlab = "z", ylab = expression(f(z)), xaxt = "n")
abline(h = 0)
axis(side = 1, at = c(-1.96, 1.96), labels = c(expression(-z[alpha/2]), expression(z[alpha/2])))
polygon(c(-4,seq(-4,-1.96,0.01),-1.96), c(0,dnorm(seq(-4,-1.96,0.01)), 0), col = "grey70", border = NA)
polygon(c(4,seq(4,1.96,-0.01),1.96), c(0,dnorm(seq(4,1.96,-0.01)), 0), col = "grey70", border = NA)
abline(v = 1.96, lty = 2, col = "grey60")
abline(v = -1.96, lty = 2, col = "grey60")
## text(x = 1.94, y = -0.04, "c", pos = 4)
## text(x = -1.94, y = -0.04, "-c", pos = 2)
text(x = 0, y = dnorm(0)/2, expression(1-alpha))
text(x = 2.2, y = 0.05, expression(alpha/2), pos = 4)
text(x = -2.2, y = 0.05, expression(alpha/2), pos = 2)
```


How can we generalize this to $1-\alpha$ confidence intervals? For a standard normal rv, $Z$, we know that
$$ 
\P(-z_{\alpha/2} \leq Z \leq z_{\alpha/2}) = 1-\alpha
$$
which implies that we can obtain a $1-\alpha$ asymptotic confidence intervals by using the interval $[L, U]$, where
$$ 
L = \widehat{\theta}_{n} - z_{\alpha/2} \widehat{\se}[\widehat{\theta}_{n}], \quad U = \widehat{\theta}_{n} + z_{\alpha/2} \widehat{\se}[\widehat{\theta}_{n}]. 
$$
This is sometimes shortened to $\widehat{\theta}_n \pm z_{\alpha/2} \widehat{\se}[\widehat{\theta}_{n}]$. Remember that we can obtain the values of $z_{\alpha/2}$ easily from R:

```{r}
## alpha = 0.1 for 90% CI
qnorm(0.1 / 2, lower.tail = FALSE)
```

As a concrete example, then, we could derive a 90% asymptotic confidence interval for the sample mean as 
$$
\left[\Xbar_{n} - 1.64 \frac{\widehat{\sigma}}{\sqrt{n}}, \Xbar_{n} + 1.64 \frac{\widehat{\sigma}}{\sqrt{n}}\right]
$$


## Interpreting confidence intervals

Remember that the interpretation of confidence is how the random interval performs over repeated samples. A valid 95% confidence interval is a random interval that will contain the true value in 95% of samples. Simulating repeated samples can help to clarify this a bit. 


::: {#exm-cis}

Suppose that we are taking samples of size $n=500$ of random variables where $X_i \sim \N(1, 10)$, and we want to estimate the population mean $\E[X] = 1$. To do so, we repeat the following steps:

1. Draw a sample of $n=500$ from $\N(1, 10)$. 
2. Calculate the 95% confidence interval sample mean $\Xbar_n \pm 1.96\widehat{\sigma}/\sqrt{n}$. 
3. Plot the intervals along the x-axis and color them blue if they contain the truth (1) and red if they do not. 

@fig-ci-sim shows 100 iterations of these steps. Here we see that, as expected, the large majority of calculated CIs contain the true value. In fact, exactly 5 of the random samples produce intervals that fail to include 1, which is an exact coverage rate of 95%. Of course, this is just one simulation and a different set of 100 random samples might have produced a slightly different coverage rate. The guarantee of the 95% confidence intervals is that if we were to continue to take these repeated samples the long-run frequency of intervals covering the truth would approach 0.95. 

```{r}
#| label: "fig-ci-sim"
#| echo: false
#| fig.cap: "95% confidence intervals from 100 random samples. Intervals are blue if they contain the truth and red if they do not."
set.seed(02143)
sims<- 10000
cover <- rep(0, times = sims)
low.bound <- up.bound <- rep(NA, times = sims)
for(i in 1:sims){
    draws <- rnorm(500, mean = 1, sd = sqrt(10))
    low.bound[i] <- mean(draws) - sd(draws) / sqrt(500) * 1.96
    up.bound[i] <- mean(draws) + sd(draws) / sqrt(500) * 1.96
    if (low.bound[i] < 1 & up.bound[i] > 1) {
        cover[i] <- 1
    }
}
cols <- ifelse(cover == 1, "dodgerblue", "indianred")
plot(x = NULL, y = NULL, xlim = c(1,100), ylim = c(0.5, 1.5), bty = "n",
     xlab = "Trial", ylab = "Estimate", xaxt = "n", las = 1)
abline(h = 1, col = "grey", lwd = 3)
segments(x0 = 1:100, y0 = low.bound[1:100], y1 = up.bound[1:100],
         col = cols[1:100], lwd = 3)
```


:::

## Exact confidence intervals under normal data

In the last chapter, we saw that we could obtain tests with the correct or exact size if we make the parametric modeling assumption that the data are normally distributed. We can take a similar approach to show that, under normal data, a $100(1-\alpha)$ confidence interval for the sample mean will have coverage exactly equal to $1-\alpha$.In practice, we rarely believe this modeling assumption, but for historical reasons this is the default of many statistical software packages, and it works well in practice as an ad hoc adjustment in finite samples. 

Assume that the data are distributed iid $X_i \sim N(\mu, \sigma^2)$. Under this assumption, we know that the normalized chance errors will be distributed as
$$
\frac{\Xbar_n - \mu}{\sigma / \sqrt{n}} \sim N(0, 1)
$$
at any sample size. With asymptotically normal estimators, we relied on Slutsky's Theorem to allow us to replace the unknown population variance $\sigma^2$ with its consistent estimator $S^2_n$. In small samples, we cannot rely on that, but with normal data, we know that using the sample variance estimator leads to a Student's t-distribution,
$$
\frac{\Xbar_n - \mu}{S_n / \sqrt{n}} \sim t_{n-1}.
$$
In the continusing abuse of notation, let $t_{\nu,\alpha}$ be the $\alpha$ quantile of a Student's t-distribution with $\nu$ degrees of freedom so that $\P(T_{\nu} \leq t_{\nu, \alpha}) = \alpha$. Then, for a 95% confidence interval with $n=50$, the large normalize shocks would be $t_{49, 0.975} = 2.01$ on the positive side and $t_{49, 0.025} = -2.01$ on the negative side. Plugging these in above and solving for the upper and lower bounds gives
$$
[L, U] = \left[\Xbar_{50} - 2.01 \times \frac{S_{50}}{\sqrt{50},} \Xbar_{50} + 2.01 \times \frac{S_{50}}{\sqrt{50}} \right].
$$
More generally, the $100(1-\alpha)$ confidence interval for the sample mean under normal data with $n$ iid draws would be
$$
[L, U] = \left[\Xbar_{n} - t_{n-1,1-\alpha/2} \times \frac{S_{n}}{\sqrt{n}}, \Xbar_{n} + t_{n-1,1-\alpha/2} \times \frac{S_{n}}{\sqrt{n}} \right].
$$


## Confidence intervals and hypothesis tests


At first glance, we may seem sloppy in using $\alpha$ in the derivation of a $1 - \alpha$ confidence interval in this chapter and in an $\alpha$-level test in the last chapter. In reality, we were simply foreshadowing the deep connection between the two: every $1-\alpha$ confidence interval contains all null hypotheses that we **would not reject** with an $\alpha$-level test. 

This connection is easiest to see with a asymptotically normal estimator, $\widehat{\theta}_n$. Consider the hypothesis test of 
$$ 
H_0: \theta = \theta_0 \quad \text{vs.}\quad H_1: \theta \neq \theta_0,
$$
using the test statistic,
$$ 
T = \frac{\widehat{\theta}_{n} - \theta_{0}}{\widehat{\se}[\widehat{\theta}_{n}]}. 
$$
As we discussed in the last chapter, an $\alpha = 0.05$ test would reject this null when $|T| > 1.96$, or when 
$$ 
|\widehat{\theta}_{n} - \theta_{0}| > 1.96 \widehat{\se}[\widehat{\theta}_{n}]. 
$$
Notice that will be true when 
$$ 
\theta_{0} < \widehat{\theta}_{n} - 1.96\widehat{\se}[\widehat{\theta}_{n}]\quad \text{ or }\quad \widehat{\theta}_{n} + 1.96\widehat{\se}[\widehat{\theta}_{n}] < \theta_{0}
$$
or, equivalently, that null hypothesis is outside of the 95% confidence interval, $$\theta_0 \notin \left[\widehat{\theta}_{n} - 1.96\widehat{\se}[\widehat{\theta}_{n}], \widehat{\theta}_{n} + 1.96\widehat{\se}[\widehat{\theta}_{n}]\right].$$ 
Of course, our choice of the null hypothesis was arbitrary, which means that any null hypothesis that is outside the 95% confidence interval would be rejected by a $\alpha = 0.05$ level test of that null. And any null hypothesis inside the confidence interval is a null hypothesis that we would not reject. 

This relationship holds more broadly. Any $1-\alpha$ confidence interval contains all possible parameter values that would not be rejected as the null hypothesis of an $\alpha$-level hypothesis test. This can be really useful for two reasons:

1. We can quickly determine if a null hypothesis is rejected at some level by inspecting if it falls in a confidence interval. 
2. There are situations where determining a confidence interval might be difficult, but performing a hypothesis test is straightforward. Then, we can find the rejection region for the test and determine what null hypotheses would not be rejected at level $\alpha$ to formulate the $1-\alpha$ confidence interval. This process is called **inverting a test**. One important application of this method is for formulating confidence intervals for treatment effects based on randomization inference in the finite population analysis of experiments. 

<!--# TODO: add example of inverting a test?-->

## Resampling approaches to confidence intervals

The entire approach to constructing confidence intervals depends on having an estimator for the standard error. For simple cases like the sample mean, we have simple and straightforward estimators for the standard error, such as $\widehat{\se} = \widehat{\sigma}/\sqrt{n}$. For some more complicated quantities of interest, such as $\mu^2$ or $\log(\mu)$ with estimators $\Xbar_{n}^2$ and $\log(\Xbar_{n})$, we saw in @sec-delta-method that the delta method can provide estimators for the standard error. 

What about quantities of interest and estimators that are not nice differentiable function, such as the median? How do we obtain the standard error in those cases? We could turn to our local department of mathematics for aid, but why think harder when we can get a computer to think for us. In the 20th century, statisticians figured out that we can approximate standard errors using computational approaches.  

All of the computational approaches to confidence intervals (and hypothesis tests) rely on a common idea: resampling. Our goal with statistical inference is to learn about the entire distribution of the estimator, but we only have a single sample from the population distribution. Ideally, we could repeatedly sample from the population (as we have done in many of our simulation exercises in this book). Imagine, we could get $R$ random samples from the population and obtain estimates from each:
$$
\left(\widehat{\theta}_{1}, \widehat{\theta}_{2}, \ldots, \widehat{\theta}_{R}\right) 
$$
These estimates represent $R$ draws from the sampling distribution of $\widehat{\theta}$. Recall we only get one of these draws, but what if we had them all? Then we could estimate the standard error with the empirical standard deviation of these estimates:
$$
\sqrt{\frac{1}{R} \sum_{r=1}^R \left(\widehat{\theta}_{r} - \overline{\theta}_{R}\right)^2},
$$
where $\overline{\theta}_{R} = (1/R) \sum_{r=1}^R \widehat{\theta}_{r}$ is the average of the $R$ samples. 

Resampling approaches to inference generate additional samples from the single original sample (thus leading to the name *re*sampling). The key difference between different methods is how we do the resampling. 

### The jackknife

The jackknife is a resampling technique that generates additional samples by removing observations. Let $\widehat{\theta}_{(-i)}$ be the estimate when omitting the $i$th observation from the sample and let $\overline{\theta}_{n} = (1/n) \sum_{i=1}^n \widehat{\theta}_{(-i)}$ be the average of these leave-one-out estimates. Under some conditions, the variation of $\widehat{\theta}_{(-i)}$ around $\overline{\theta}_{n}$ should be similar to the variation of $\widehat{\theta}_{n}$ around $\theta$. This fact motivates the jackknife variance estimator, 
$$
V_{jk} = \frac{n-1}{n} \sum_{i=1}^n \left(\widehat{\theta}_{(-i)} - \overline{\theta}_{n}\right)^2,
$$
where the $n-1$ in the numerator accounts for the reduced variability of the jackknife samples compared to "fresh" samples. Indeed, two jackknife samples share $n-2$ common values.

Under iid sampling and some regularity conditions (namely that the estimator is a smooth function of the data), the jackknife variance estimator will be consistent in the sense that 
$$
V_{jk} / \V[\widehat{\theta}_n] \inprob 1
$$

With this variance estimator in hand, one can derive confidence intervals in the usual way, dropping in the jackknife variance estimator for the usual one,
$$
CI_{\alpha, jk} = \left[\widehat{\theta}_n - z_{\alpha/2}\sqrt{\frac{V_{jk}}{n}}, \widehat{\theta}_n + z_{\alpha/2}\sqrt{\frac{V_{jk}}{n}} \right] 
$$


We can see why we use the $n-1$ adjustment, we can focus on the easiest case: the sample mean. Define the leave-one-out sample mean as 
$$
\Xbar_{(i)} = \frac{1}{n-1} \sum_{j \neq i} X_j,
$$
and notice that we can easily write this in terms of the full sample mean,
$$
\Xbar_{(i)} = \frac{1}{n-1}\left(\sum_{j=1} ^n X_j - X_i \right) = \frac{1}{n-1}\left(n\Xbar_{n} - X_i \right). 
$$

From this expression, we can also show that the mean of the jackknife estimates is the same as the full sample mean:
$$
\begin{aligned}
\Xbar_{jk} = \frac{1}{n} \sum_{i=1}^n \Xbar_{(i)} &= \frac{1}{n(n-1)} \sum_{i=1}^n n\Xbar_{n} - X_i \\
&= \frac{n}{n-1} \Xbar_n - \frac{1}{n-1} \Xbar_n \\
&= \Xbar_n
\end{aligned}
$$

From these two facts, we can see that the jackknife deviations rescaled version of the original deviations, 
$$
(\Xbar_{(i)} - \Xbar_{jk}) = \frac{n}{n-1} \Xbar_n - \frac{1}{n-1} X_i - \Xbar_n = \frac{1}{n-1}(\Xbar_n - X_i)
$$
This key fact justifies the jackknife variance estimator since, 
$$
\begin{aligned}
V_{jk} &= \frac{n-1}{n} \sum_{i=1}^n (\Xbar_{(i)} - \Xbar_{jk})^2 \\
&= \frac{n-1}{n(n-1)^2} \sum_{i=1}^n (X_i -\Xbar_n)^2 \\
&=\frac{\widehat{\sigma}^2}{n},
\end{aligned}
$$
which is the typical estimator for the sampling variance of the sample mean. 

These derivations show how the jackknife variance estimator for the sample mean is actually identical to the usual sampling variance estimator. Obviously, this fact means that the jackknife does not have a huge benefit for the sample mean itself. But the basic intuition---that the jackknife deviations are similar to a $(n-1)^{-1}$ scaled version of full-sample deviation---generalizes to other estimators that are smooth functions of the data where the jackknife could be more fruitfully applied. 

The smoothness requirements of the jackknife are actually more restrictive than they appear. In particular, it rules out using the jackknife for sample quantiles like the sample median. There are modifications of the jackknife, such as the **delete-d jackknife** (where we drop $d>1$ observations at a time) that have better properties for these types of estimators, but there are also alternative resampling approaches that work well in more general settings. 

### The nonparametric bootstrap

Another way to generate additional artificial samples is to randomly sample from the data itself. The **nonparametric bootstrap** is a resampling technique that generates a distribution of estimates by repeatedly sampling the data with replacement. In particular, we can define the empirical distribution of the data as $F_n$, which places $1/n$ probability on each observed data point. This distribution has a cdf of 
$$
F_n(x) = (1/n) \sum_{i=1}^n \mathbb{1}(X_i \leq x).
$$
Let $X^{\ast}_1, X^{\ast}_2, \ldots, X^{\ast}_n \sim F_n$ be an iid sample (sample with replacement) from this empirical distribution. We call this a **bootstrap sample** from the notion that we are pulling ourselves up by our own bootstraps (or our own sample data). We can then compute our estimator on this bootstrap sample, $\widehat{\theta}^* = \theta(X^{\ast}_1, X^{\ast}_2, \ldots, X^{\ast}_n)$. 

Intuitively, we are treating the sample data as if it were the population. With this pseudo-population, we can repeated sample as many bootstrap samples as our hearts desire (and our computers can handle). In each bootstrap sample, we can compute our estimator, giving us a large number of bootstrap draws of the estimator. 

How can use these bootstrap estimates to derive a confidence interval? One way to think about confidence intervals is that they encode what values of the true parameter are consistent with a "plausible" range of chance errors. In the asymptotically normal case above, we might imagine asking: what if our estimate had a large positive chance error, so that $\widehat{\theta}_n - \theta$ was larger than 97.5\% of errors. This would imply that $\widehat{\theta}_n - \theta = 1.96 \se[\widehat{\theta}_n]$, and the implied true value in this case would be $\theta = \widehat{\theta}_n - 1.96 \se[\widehat{\theta}_n]$

::: {#def-bootstrap-variance}

## Basic bootstrap confidence interval

1. Draw a sample with replacement of size $n$ from $F_n$ and apply the estimator to this sample, $\widehat{\theta}_r$.
2. Repeat Step 1 $R$ time to obtain $\widehat{\theta}_1, \widehat{\theta}_2, \ldots, \widehat{\theta}_R$. 
3. Calculate the $\alpha/2$ and $1-\alpha/2$ quantiles of the distribution of $(\widehat{\theta}_1 - \widehat{\theta}_n, \ldots, \widehat{\theta}_R - \widehat{\theta}_n)$, which are the bootstrap estimates minus the full sample estimate, and define these as $t_{\alpha/2}$ and $t_{1-\alpha/2}$. 

4. Construct a $(1-\alpha)$ confidence interval as 
$$
CI_{1-\alpha, boot} = \left[\widehat{\theta}_n - t_{1-\alpha/2}, \widehat{\theta}_{n} - t_{\alpha/2}\right]
$$ {#eq-bootstrap-cis}

:::


In the construction of the bootstrap confidence intervals, it may be confusing that we subtract both quantiles from the estimate, rather than add one and subtract one. But recall the derivation of confidence intervals in @sec-deriving-cis as using the distribution of chance errors. The bootstrap approach leverages the  "bootstrap deviations" such as $\widehat{\theta}_r - \widehat{\theta}_n$ to approximate the distribution of $\widehat{\theta}_n - \theta$. The quantile $t_{1-\alpha/2}$ ($0.975$ for $\alpha = 0.05$) of these deviations will represent a large positive chance deviation, meaning that the lower bound $\widehat{\theta}_n - t_{1-\alpha/2}$ of the confidence interval will be below $\widehat{\theta}_n$. The quantile $t_{\alpha/2}$, on the other hand, ($0.025$ for $\alpha = 0.05$) will be a large negative chance error, meaning the upper bound $\widehat{\theta}_n - t_{\alpha/2}$ will be larger than $\widehat{\theta}$.  

Just like for the asymptotically normal estimators in @sec-sec-deriving-cis, we can back out the implied true value from these deviations, leading to the bootstrap confidence intervals formula in @eq-bootstrap-cis. 


